# The sf package for spatial vector data

**Learning objectives:**

- learn what simple features are and get know them in R
- read, write, decompose and compose an `sf` object
- get the gist of making a static and an interactive map (see ch. 5)
- work with `sf` objects: subsetting, spatial operations, logical operations, joining

## Simple features (1) {-}

- [Simple features](https://en.wikipedia.org/wiki/Simple_Features) ([simple feature access](https://www.ogc.org/standard/sfa/)): formal standard (ISO 19125-1:2004) to describe how **objects in the real world** can be represented in computers, with emphasis on the **spatial geometry**
- All geometries are composed of points. Points are coordinates in a 2-, 3- or 4-dimensional space.

(Source: [Vignette 1](https://r-spatial.github.io/sf/articles/sf1.html) of **sf** package)

## Simple features (2) {-}

Seven most common types:

| type | description                                       
| ---- | -------------------------------------------------- 
| `POINT` | single point
| `LINESTRING` | sequence of points connected by straight, non-self intersecting line pieces
| `POLYGON` | sequence of points form a closed, non-self intersecting ring; zero or more subsequent rings denote holes
| `MULTIPOINT` | set of points
| `MULTILINESTRING` | set of linestrings
| `MULTIPOLYGON` | set of polygons
| `GEOMETRYCOLLECTION` | set of geometries of any type except `GEOMETRYCOLLECTION`

(Based on [vignette 1](https://r-spatial.github.io/sf/articles/sf1.html) of **sf** package)

## Simple features in R (1) {-}

Package **sf**: to read, manipulate and write simple features in R

```{r message=FALSE}
# install.packages("sf")
library(sf)
filepath <- system.file("shape/nc.shp", package = "sf")
```

Reading a geospatial data file:

```{r}
# nc <- st_read(filepath, quiet = TRUE, as_tibble = TRUE)
nc <- read_sf(filepath)
```

Read the documentation of this dataset with `?nc` and on <https://jakubnowosad.com/spData/reference/nc.sids.html>.

## Simple features in R (2) {-}

```{r}
nc
```

## Simple features in R (3) {-}

```{r}
library(dplyr, warn.conflicts = FALSE)
glimpse(nc)
```

## Simple features in R (4) {-}

Extracting the geometry column:

```{r collapse = TRUE}
nc_geom <- st_geometry(nc)
nc_geom
```

```{r}
all.equal(nc_geom, nc$geometry)
```

## Simple features in R (5) {-}

```{r}
class(nc)
class(nc_geom)
```

- Class `sf`: data frame:
  - rows = simple features
  - columns = attributes and a 'sticky' geometry column
- Class `sfc`: the geometry column, i.e. a list column.
It is a **list** of the geometries of the simple features.

```{r}
nc_geom[[8]] # eighth geometry
class(nc_geom[[8]])
```

- Class `sfg`: the geometry of a single feature

## Simple features in R (6) {-}

Dropping the geometry column:

```{r}
st_drop_geometry(nc)
```

## Simple features in R (7) {-}

Returning the CRS (coordinate reference system):

```{r}
nc_crs <- st_crs(nc)
```

From `?st_crs`:

> the `$` method for `crs` objects retrieves named elements using the GDAL interface; named elements include `"SemiMajor"`, `"SemiMinor"`, `"InvFlattening"`, `"IsGeographic"`, `"units_gdal"`, `"IsVertical"`, `"WktPretty"`, `"Wkt"`, `"Name"`, `"proj4string"`, `"epsg"`, `"yx"`, `"ud_unit"`, and `"axes"` (this may be subject to changes in future GDAL versions).

## Simple features in R (8) {-}

```{r}
nc_crs$Name
nc_crs$epsg
```

## Simple features in R (9) {-}

Formal CRS definition is given by the WKT string (also given by `print(nc_crs)`):

```{r}
nc_crs$wkt |> cat()
```


## Simple features in R (10) {-}

Return point coordinates contained in a `sf` object:

```{r}
st_coordinates(nc) |> class()
st_coordinates(nc) |> head()
st_coordinates(nc) |> tail()
```

From `?st_coordinates`:

> [...] for `MULTIPOLYGON` `L1` refers to the main ring or holes, `L2` to the ring id in the `MULTIPOLYGON`, and `L3` to the simple feature.

## Static maps with ggplot2 (1) {-}

```{r out.width='100%'}
library(ggplot2)
ggplot(nc) + geom_sf(fill = "mistyrose") + theme_bw()
```

## Static maps with ggplot2 (2) {-}

```{r warning=FALSE}
ggplot(nc) + 
  geom_sf() + 
  geom_sf_label(aes(label = NAME)) +
  coord_sf(xlim = c(-81, -79), ylim = c(35, 36)) +
  theme_bw()
```

## Interactive maps with mapview (1) {-}

`st_sample()` samples points on spatial features, and returns an `sfc` class

However **[mapview](https://r-spatial.github.io/mapview)** needs an `sf` object.
To upgrade `sfc` to `sf` we use `st_as_sf()`.

```{r include=FALSE}
set.seed(20240203)
```

```{r}
nc_points <- st_sample(nc, size = 10) |> st_as_sf() 
```

## Interactive maps with mapview (2) {-}

```{r message=FALSE, warning=FALSE, out.width='100%'}
library(mapview)
mapview(nc, col.regions = "navajowhite3", label = "NAME") + 
  mapview(nc_points, col.regions = "red", cex = 5)
```

## Another `st_coordinates()` example {-}

```{r}
st_geometry_type(nc_points)
```

```{r}
st_coordinates(nc_points) |> head()
```

## Writing an sf object to file {-}

Use `st_write()`:

```{r}
tempfile_geojson <- tempfile(fileext = ".geojson")
st_write(nc_points, tempfile_geojson)
```

Read back in with:

```{r eval=FALSE}
st_read(tempfile_geojson)
```

## Subsetting simple features (1) {-}

Based on the **sf** documentation:

```
## S3 method for class 'sf'
x[i, j, ..., drop = FALSE, op = st_intersects]
```

- `i`: row selection (as in data frames), or a `sf` object to work with the `op` argument
- `j`: column selection (as in data frames)
- `drop`:  default `FALSE`; if `TRUE` drop the geometry column (= return data frame)
- `op`: geometrical binary predicate function to apply when `i` is a simple feature object

So we can subset using non-spatial criteria (row and column selection) **or** spatial criteria (another `sf` object).

## Subsetting simple features (2) {-}

Non-spatial criteria:

```{r}
nc[1, ] # first row
```

## Subsetting simple features (3) {-}

Non-spatial criteria -- continued:

```{r}
nc[nc$NAME == "Ashe", ] # row with NAME "Ashe"
```

## Subsetting simple features (4) {-}

Non-spatial criteria -- continued:

```{r}
nc[1, "NWBIR74"] # first row, column with name NWBIR74
nc[1, "NWBIR74", drop = TRUE] # drop geometry
```

## Subsetting simple features (5) {-}

Non-spatial criteria -- continued:

```{r out.width='100%'}
nc[!(nc$FIPS %in% c("37125", "37051")), ] |> 
  ggplot() + 
  geom_sf(aes(fill = SID79)) +
  theme_bw()
```

## Subsetting simple features (6) {-}

Spatial criteria:

```{r}
nc_subset <- nc[nc_points, ]
nc_subset
```

## Subsetting simple features (7) {-}

```{r out.width='100%'}
ggplot() + 
  geom_sf(data = nc_subset) + 
  geom_sf(data = nc_points) + 
  theme_bw()
```

## Generate sf objects (1) {-}

First let's start at a higher level: we already have a data object.

Conversion can then be done with `st_as_sf()`.

Suppose we have a data frame with point coordinates, and we know that the CRS is `EPSG:4326`.

```{r}
d <- tibble(
  place = c("London", "Paris", "Madrid", "Rome"),
  long = c(-0.118092, 2.349014, -3.703339, 12.496366),
  lat = c(51.509865, 48.864716, 40.416729, 41.902782),
  value = c(200, 300, 400, 500))
d
```

## Generate sf objects (2) {-}

```{r}
dsf <- st_as_sf(d, coords = c("long", "lat"), crs = 4326)
dsf
```

## Generate sf objects (3) {-}

We can also start from scratch:

- create the geometries from coordinate vectors or matrices with `st_point()`, `st_linestring()`, `st_polygon()`, `st_multipolygon()` etc.
- combine the geometries into an `sfc` object with `st_sfc()`, which also handles the CRS
- create an attribute dataframe
- combine data frame and the geometry column with `st_sf()`

## Generate sf objects (4) {-}

```{r}
# Single point (point as a vector)
p1_sfg <- st_point(c(2, 2))
p2_sfg <- st_point(c(2.5, 3))
```

```{r}
p1_sfg
p2_sfg
```

## Generate sf objects (5) {-}

```{r}
# Set of points (points as a matrix)
p <- rbind(c(6, 2), c(6.1, 2.6), c(6.8, 2.5),
           c(6.2, 1.5), c(6.8, 1.8))
p
mp_sfg <- st_multipoint(p)
mp_sfg
```

## Generate sf objects (6) {-}

A polygon according to the **sf** vignette:

- a geometry with a positive area (two-dimensional)
- sequence of points form a closed, non-self intersecting ring
  - the first ring denotes the exterior ring
  - zero or more subsequent rings denote holes in this exterior ring

## Generate sf objects (7) {-}

```{r}
# exterior ring
p1 <- rbind(c(10, 0), c(11, 0), c(13, 2),
            c(12, 4), c(11, 4), c(10, 0))
p1
# hole
p2 <- rbind(c(11, 1), c(11, 2), c(12, 2), c(11, 1))
p2
# polygon including hole
pol_sfg <- st_polygon(list(p1, p2))
pol_sfg
```

## Generate sf objects (8) {-}

```{r collapse=TRUE}
p_sfc <- st_sfc(p1_sfg, p2_sfg, mp_sfg, pol_sfg)
p_sfc
```

## Generate sf objects (9) {-}

```{r}
df <- tibble(v1 = c("A", "B", "C", "D"))
df
p_sf <- st_sf(df, geometry = p_sfc)
p_sf
```

## Generate sf objects (10) {-}

```{r out.width='100%'}
ggplot(p_sf) + 
  geom_sf(aes(fill = v1), size = 3, shape = 21) + 
  theme_bw()
```


## Manipulating sf objects (1) {-}

CRS related:

- `st_crs()<-` sets a CRS
- `st_transform()` transforms data to another CRS (see previous chapter)

Spatial operations, e.g.:

- `st_union()` combines several sf objects into one
- `st_simplify()` simplifies a sf object

## Manipulating sf objects (2) {-}

```{r collapse=TRUE}
nc_union <- st_union(nc)
nc_union
```


```{r}
class(nc_union)
```

## Manipulating sf objects (3) {-}

`ggplot()` also supports `sfc`:

```{r out.width='100%'}
ggplot(nc_union) + geom_sf() + theme_bw()
```

## Manipulating sf objects (5) {-}

```{r out.width='100%'}
ggplot(st_simplify(nc, dTolerance = 5e3)) + geom_sf() + theme_bw()
```

## Manipulating sf objects (4) {-}

```{r out.width='100%'}
ggplot(st_simplify(nc, dTolerance = 10e3)) + geom_sf() + theme_bw()
```

## Manipulating sf objects (6) {-}

```{r out.width='100%'}
ggplot(st_simplify(nc, dTolerance = 15e3)) + geom_sf() + theme_bw()
```

## Manipulating sf objects (7) {-}

Some more spatial operations are shown in <https://r.geocompx.org/geometry-operations#fig:venn-clip>.

## Binary logical operations (1) {-}

Partly [based on vignette 3](https://r-spatial.github.io/sf/articles/sf3.html#binary-logical-operations) from **sf**.

First create some objects to play with.

- `x`: 3 touching polygons
- `y`: 4 overlapping polygons

```{r}
b0 <- st_polygon(list(rbind(c(-1, -1), c(1, -1), c(1, 1), c(-1, 1), c(-1, -1))))
b1 <- b0 + 2
b2 <- b0 + c(-0.2, 2)
x <- st_sfc(b0, b1, b2)
a0 <- b0 * 0.8
a1 <- a0 * 0.5 + c(2, 0.7)
a2 <- a0 + 1
a3 <- b0 * 0.5 + c(2, -0.5)
y <- st_sfc(a0, a1, a2, a3)
plot(x, border = 'red')
plot(y, border = 'green', add = TRUE)
```

## Binary logical operations (2) {-}

Binary logical operations (binary predicate functions) describe the topological relationship between **a pair of simple features**.

Applied to two `sf` objects, they return:

- either a _sparse_ matrix (`sgbp` class: sparse geometry binary predicate),
- or a _dense_ matrix.

```{r}
sparse <- st_intersects(x, y)
sparse
class(sparse)
str(sparse, give.attr = FALSE)
```

## Binary logical operations (3) {-}

```{r}
dense <- st_intersects(x, y, sparse = FALSE)
dense
class(dense)
str(dense)
```

## Binary logical operations (4) {-}

An overview of binary logical operations is illustrated in <https://r.geocompx.org/spatial-operations#fig:relations>.

## Binary logical operations (5) {-}

With binary predicate functions like `st_intersects()`, one can collect information about the topological relationship _without making a spatially joined `sf` object_ (with `sf::st_join()`).

**Example 1**: counting points in polygons and adding the result to a copy of `nc`.

```{r include=FALSE}
set.seed(20230203)
```

```{r}
nc_points_2 <- st_sample(nc, size = 100)
```

## Binary logical operations (6) {-}

```{r out.width='100%'}
ggplot() + geom_sf(data = nc) + geom_sf(data = nc_points_2) + theme_bw()
```

## Binary logical operations (7) {-}

The order of the two arguments of `st_intersects()` is important!

```{r}
inter <- st_intersects(nc, nc_points_2)
inter
lengths(inter)
```

## Binary logical operations (8) {-}

Add point count to each polygon:

```{r}
nc2 <- nc
nc2$count <- lengths(inter)
glimpse(nc2)
```

## Binary logical operations (9) {-}

```{r out.width='100%'}
ggplot(nc2) + geom_sf(aes(fill = count)) + theme_bw()
```

## Binary logical operations (10) {-}

**Example 2**: identifying the `nc` polygons that contain points from `nc_points`.

Notice the order of the arguments in `st_intersects()`!

```{r}
inter <- st_intersects(nc_points, nc)
inter
unlist(inter)
```

## Binary logical operations (11) {-}

```{r}
counties_with_point <- nc[unlist(inter), ]
counties_with_point$NAME
```

```{r}
cbind(nc_points, areaname = counties_with_point$NAME)
```


## Joining `sf` object with data (1) {-}

This can be done with the `*_join()` functions from **dplyr**.

Example [based on vignette 4](https://r-spatial.github.io/sf/articles/sf4.html#joining-two-feature-sets-based-on-attributes) from **sf**.

```{r}
x <- st_sf(a = 1:2, geom = st_sfc(st_point(c(0, 0)), st_point(c(1, 1))))
y <- data.frame(a = 2:3, letter = letters[11:12])
x
y
```

## Joining `sf` object with data (2) {-}

```{r}
inner_join(x, y, by = "a")
```

## Joining `sf` object with data (3) {-}

```{r}
left_join(x, y, by = "a")
```

## Joining `sf` object with data (4) {-}

```{r}
semi_join(x, y, by = "a")
```


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
