# Spatial neighbourhood matrices

**Learning objectives:**

- understand what spatial neighbours are
- know how spatial neighbours can be defined
- create and plot a neighbours list
- use a neighbours list to create a spatial neighbourhood matrix

## Areal data {-}

- This is the first chapter of the part 'areal data'.

- > In areal or lattice data, the domain D  is a fixed countable collection of (regular or irregular) areal units at which variables are observed.

- Areal data usually arise when a number of events corresponding to some variable of interest are aggregated in areas.

## Spatial neighbourhood {-}

- It represents which areas are close to one another (polygons, points)

  - In this chapter, we won't use attribute variables, only the geometries

## Spatial neighbourhood {-}

- It will help to assess spatial autocorrelation with areal data
  - to do that, areas must be spatially connected by weights: in a **spatial neighbourhood matrix**
  - to obtain a neighbourhood matrix, one needs to define the neighbours of each area: the **neighbours list**

## Spatial neighbourhood in R {-}

- Package **spdep**: <https://r-spatial.github.io/spdep>

```{r message=FALSE}
library(sf)
library(spdep)
library(ggplot2)
```

## Read example data {-}

```{r}
map <- read_sf(system.file("shapes/columbus.shp",
                           package = "spData"), quiet = TRUE)
map
```

## Example data {-}

From `?spData::columbus`:

> The columbus data frame has 49 rows and 22 columns

> Unit of analysis: 49 neighbourhoods in Columbus, OH, 1980 data

## Example data {-}

```{r}
# we won't need attributes:
map_geom <- st_geometry(map)
ggplot(map_geom) + geom_sf() + theme_bw()
```

## Spatial neighbourhood {-}

Remind:

- **spatial neighbourhood matrix**: connects areas by weights
- to obtain it, one needs the **neighbours list**: defines the neighbours of each area

## Spatial neighbours  {-}

The concept of a neighbour is **binary** (0 / 1)!

Area 2 **is** a spatial neighbour of area 1, or it is **not**.

## Spatial neighbours list  {-}

- A **neighbours list** (**`nb`** class) is a kind of sparse matrix: a list that gives the indices of neighbours for each area in turn.
  - e.g. the first 6 elements give the neighbour indices of the first 6 geometries of the input layer:
  
```r
[[1]]
[1] 2 3
[[2]]
[1] 1 3 4
[[3]]
[1] 1 2 4 5
[[4]]
[1] 2 3 5 8
[[5]]
[1]  3  4  6  8  9 11 15 16
[[6]]
[1] 5 9
```

## Defining who is a neighbour and who isn't {-}

- **contiguity criteria** -- this needs _polygons_:
  - the areas that share at least a common vertex (type **Queen**)
  - the areas that share a common border (type **Rook**)

- **distance criteria** -- this needs _points_ (e.g. polygon centroids):
  - the areas that are **within some distance** apart (lower and upper bounds)
  - the areas that are **among the $k$ nearest** to an area (asymmetric relationship)

## Creating a neighbours list ('nb') from geometries {-}

- contiguity based:
  - `poly2nb(<polygons>, queen = TRUE)` (default)
  - `poly2nb(<polygons>, queen = FALSE)`
  
- distance based:
  - `dnearneigh(<points>, d1, d2)`
  - `knn2nb(<matrix of nearest neighbours>)`

## Neighbours list: type Queen contiguity {-}

```{r}
nb1 <- poly2nb(map_geom, queen = TRUE)
nb1
head(nb1)
```

## Neighbours list: type Rook contiguity {-}

```{r}
nb2 <- poly2nb(map_geom, queen = FALSE)
nb2
head(nb2)
```

## Plotting {-}

With `nb.plot(<neighbours list>, <sfc object>)`

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nb1, map_geom, add = TRUE)
```

## Plotting {-}

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nb2, map_geom, add = TRUE)
```

## Neighbours list based on distance bounds {-}

Creating centroids from sf polygons:

```{r collapse=TRUE}
(centroids <- st_centroid(map_geom))
```

## Neighbours list based on distance bounds {-}

```{r out.width='100%'}
ggplot() + 
  geom_sf(data = map_geom) +
  geom_sf(data = centroids) +
  theme_bw()
```

## Neighbours list based on distance bounds {-}

```{r}
nb3 <- dnearneigh(x = centroids, d1 = 0, d2 = 0.4)
head(nb3)
```

## Neighbours list based on distance bounds {-}

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nb3, map_geom, add = TRUE)
```

## Neighbours list based on $k$ nearest neighbours {-}

In two steps:

1. `knearneigh()`: create a `knn` object ('k-nearest neighbour classification')
    - it contains `nn`: a _matrix_ that defines the k nearest neighbors
1. `knn2nb()`: convert the `knn` object to a neighbours list

```{r}
knn_centroids <- knearneigh(centroids, k = 3)
class(knn_centroids)
class(knn_centroids$nn)
head(knn_centroids$nn)
```

## Neighbours list based on $k$ nearest neighbours {-}

Step 2:

```{r}
nb4 <- knn2nb(knn_centroids)
head(nb4)
```

## Neighbours list based on $k$ nearest neighbours {-}

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nb4, map_geom, add = TRUE)
```

## Creating higher order neighbours lists {-}

Starting from an existing neighbours list, one can redefine neighbours using a lag:

- lag = 2: neighbours are 2 links apart in the original neighbours list
- lag = 3: neighbours are 3 links apart in the original neighbours list
- ...

## Creating higher order neighbours lists {-}

`nblag(<neighbours list>, maxlag =)`: to produce `maxlag` higher order neighbours lists

  - returns a list of lagged neighbours lists: element 1 for lag = 1, etc)

## Creating higher order neighbours lists {-}

```{r}
nblags <- nblag(neighbours = nb1, maxlag = 3)
class(nblags)
length(nblags)
all.equal(nb1, nblags[[1]], check.attributes = FALSE)
```

## Creating higher order neighbours lists {-}

```{r}
lapply(nblags, head, 2)
```

## Creating higher order neighbours lists {-}

Plotting the second order neighbours list:

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nblags[[2]], map_geom, add = TRUE)
```

## Creating higher order neighbours lists {-}

Plotting the third order neighbours list:

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nblags[[3]], map_geom, add = TRUE)
```

## Cumulating neighbours lists {-}

You can cumulate multiple neighbour lists to a single neighbour list:\
`nblag_cumul(<list of neighbours lists>)`

Cumulating the 1st and 2nd order neighbours lists from before:

```{r}
nblagsc <- nblag_cumul(nblags[1:2])
class(nblagsc)
head(nblagsc)
```

## Cumulating neighbours lists {-}

```{r out.width='100%'}
plot(map_geom, border = "lightgray")
plot.nb(nblagsc, map_geom, add = TRUE)
```

## Further things to do with a neighbours list {-}

- Count neighbours: `lengths(<nb>)` (or `spdep::card()`)
- Compute distances between neighbours: `nbdists(<nb>, <points>)`
- Create a spatial neighbourhoods matrix: `nb2mat(<nb>, ...)`

## Count neighbours {-}

```{r}
lengths(nb1)
```

## Compute distances between neighbours {-}

```{r}
nbdists(nb1) |> try()
```

## Compute distances between neighbours {-}

```{r}
nbdists(nb1, centroids) |> head()
```

## Neighbourhood matrix {-}

Straightforward function is the `nb2mat()` function (not in the book).

It converts the 'sparse' neighbours list to a square neighbourhood matrix of **weights**.

## Neighbourhood matrix {-}

Basic conversion from the neighbours list to a neighbourhood matrix:

```{r}
nb2mat(nb1, style = "B") |> dim()
```

## Neighbourhood matrix {-}

The basic (B) format uses its input as-is: binary!

```{r}
nb2mat(nb1, style = "B")[1:4, 1:7]
```

## Neighbourhood matrix {-}

But one can standardise, e.g. by row (W):

```{r}
nb2mat(nb1, style = "W")[1:4, 1:7] |> round(2)
```

## Neighbourhood matrix {-}

You can use `glist` argument of `nb2mat()` to replace the 0 / 1 value from the neighbours list by preset weights.

For example, calculate inverse distance weights and feed them to `nb2mat()`.

```{r}
dists <- nbdists(nb1, centroids)
head(dists)
```

## Neighbourhood matrix {-}

```{r}
ids <- lapply(dists, function(x) {1 / x})
head(ids)
```

## Neighbourhood matrix {-}

```{r}
nb2mat(nb1, glist = ids, style = "B")[1:4, 1:7]
```

## Neighbourhood matrix {-}

```{r}
nb2mat(nb1, glist = ids, style = "W")[1:4, 1:7]
```


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
