# Bayesian spatial models

**Learning objectives:**

- Understand the specification of INLA models.
- Be able to fit a bayesian model for areal data with spatial dependencies.

## Bayesian spatial models in INLA {-}

R-INLA can fit a broad class of generalized linear mixed models (GLMMs), including spatial and spatio-temporal models: those that can be expressed as latent Gaussian Markov random fields (GMRF).

In bayesian modelling, a posterior distribution is estimated for every model parameter.

INLA algorithm (Integrated Nested Laplace Approximation): fast approximation method to obtain the posterior distributions; very fast compared to MCMC approximation.

R-INLA provides several criteria to compare and evaluate models (DIC, WAIC, CPO).

## Bayesian spatial models in INLA {-}

```{r message=FALSE}
library(INLA)
library(sf)
library(spData)
library(ggplot2)
library(spdep)
```

## Bayesian spatial models in INLA {-}

The model for an observed response $Y_i$ needs:

- the likelihood distribution (exponential family, e.g. normal, poisson, binomial)
- the link function $g$
- a linear predictor $\eta_i$

## Bayesian spatial models in INLA {-}

Likelihood distribution for each observed response.

$$Y_i|\eta_i,\theta_1 \sim \pi(Y_i|\eta_i,\theta_1)$$
Example distributions:

$$Y_i \sim \mathcal{N}(\mu_i, \sigma^2_i)$$

$$Y_i \sim \mathcal{Poisson}(\mu_i)$$

$$Y_i \sim \mathcal{Bin}(n, \pi_i)$$

## Bayesian spatial models in INLA {-}

Available likelihood families:

```{r}
inla.list.models("likelihood")
```


## Bayesian spatial models in INLA {-}

The linear predictor (latent Gaussian random field) is the model of a parameter of the likelihood distribution, transformed with the link function.

E.g. using log link for parameter $\mu_i$:

$$\eta_i = ln(\mu_i) = X_i \cdot \beta + Z_i$$

E.g. using logit link for parameter $\pi_i$:

$$\eta_i = ln(\frac{\pi_i}{1 - \pi_i}) = X_i \cdot \beta + Z_i$$

## Bayesian spatial models in INLA {-}

Typically, the linear predictor contains both:

- (independent) linear covariate effects: $X_i \cdot \beta$ \
These will typically correspond to the fixed effects of frequentist GLMMs.
- other covariate effects: $Z_i$ \
E.g. non-linear effects, spatial, time & seasonal patterns, random intercepts & slope.


## Bayesian spatial models in INLA {-}

Available link functions for each likelihood family, e.g. gamma distribution:

```{r}
inla.models()$likelihood[["gamma"]]$link
```

## Bayesian spatial models in INLA {-}

Available 'latent' models to use in terms of the linear predictor:

```{r}
inla.list.models("latent")
```

## Bayesian spatial models in INLA {-}

Model specification in R-INLA:

```{r eval=FALSE}
inla(
  formula = NULL,
  family = "gaussian",
  data = NULL,
  control.compute = list(),
  control.predictor = list()
)
```

Formula of the linear predictor uses `f()` to specify latent models.

## Bayesian spatial models in INLA {-}

Package website: <https://www.r-inla.org>

Books: <https://www.r-inla.org/learnmore/books>

## Latent models referred in this chapter {-}

BYM = Besag-York-Mollié (BYM) model = the sum of:

- spatial random effect $u_i$ modelled with CAR (conditional autoregressive model; Besag model)

$$u_i|u_{-i} \sim \mathcal{N}(\overline{u_{\delta_i}}, \frac{\sigma^2_u}{n_{\delta_i}})$$

- iid (unstructured) random effects $v_i \sim \mathcal{N}(0, \sigma^2_v)$

## Latent models referred in this chapter {-}

BYM2 = different parametrisation of BYM

$$b_i = \frac{1}{\sqrt{\tau}} \cdot (\sqrt{1-\phi} \cdot v_i + \sqrt{\phi} \cdot u_i)$$

```{r eval=FALSE}
inla.doc("^besag$", section = "latent")
inla.doc("^bym$", section = "latent")
inla.doc("^bym2$", section = "latent")
```

## Case: housing prices in Boston {-}

```{r out.width='100%'}
map <- st_read(system.file("shapes/boston_tracts.shp",
                           package = "spData"), quiet = TRUE)
map$vble <- log(map$MEDV)
dim(map)
```


## Case: housing prices in Boston {-}

```{r out.width='100%'}
ggplot() + geom_sf(data = map, aes(fill = vble))
```

## Case: housing prices in Boston {-}

We will model the logarithm of the median prices using as covariates the per capita crime (`CRIM`) and the average number of rooms per dwelling (`RM`)

For spatial areas, we need an index vector to identify each area for the BYM2 latent model.

```{r eval = FALSE}
map$re_u <- 1:nrow(map)
map$re_v <- 1:nrow(map)
```

BYM2 also needs a spatial neighbourhood list formatted for INLA

```{r eval = FALSE}
nb <- poly2nb(map)
adjmat_path <- file.path(tempdir(), "map.adj")
nb2INLA(adjmat_path, nb)
g <- inla.read.graph(filename = adjmat_path)
```

```{r eval = FALSE}
formula <- vble ~ CRIM + RM + f(re_u, model = "bym2", graph = g)
```

## Case: housing prices in Boston {-}

Fitting the INLA model:

```{r eval=FALSE}
res <- inla(formula, family = "gaussian", data = map,
            control.predictor = list(compute = TRUE),
            control.compute = list(return.marginals.predictor = TRUE))
```

Control computation of fitted values (at the scale of the linear predictor):

- `control.predictor = list(compute = TRUE)`: fitted values and their posterior summary: `res$summary.fitted.values`
- `control.compute = list(return.marginals.predictor = TRUE)`: 
    compute posterior marginal distribution of the fitted values:        `res$marginals.fitted.values[[1]]`

Model object evaluation: see book.


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
