# Model-based geostatistics

**Learning objectives:**

- understand that random effects of statistical models can be a Gaussian Random Field (GRF)
- fit models with spatial correlation using R-INLA

## Aim {-}

Fit a statistical model that:

- can support various distribution families for the response variable
- accommodates fixed and random effects
- captures spatial correlation structure as a random effect
- provides spatial predictions with uncertainty measures

We will specify the spatial random effect as a Gaussian Random Field (GRF) with zero mean and a Matérn correlation.

## INLA and GRFs {-}

INLA can fit models containing a Gaussian Markov Random Field (GMRF).

It does this by solving an SPDE (stochastic partial differential equation) on the vertices of a triangulated mesh.

Generating the GMRF by solving the SPDE on the mesh can be controlled by:

- setting the smoothness parameter of the spatial autocorrelation
- requiring zero mean

## INLA and GRFs {-}

GRMF values fitted at vertices are then interpolated to locations of interest:

- fitted values at observation locations
- predicted values at prediction locations

These interpolations need a **projection matrix**: defines the relation between locations and mesh vertices by means of weights.

## Projection matrix {-}

- The rows represent the locations of interest (observations, predictions).
- The columns represent the mesh vertices.
- The values are the barycentric coordinates of the locations of interest relative to the three vertices of the triangle, the latter having a mass of 1.
- So the rowsums are 1.

The values in the projection matrix are used as _weights_ to interpolate from triangle vertices to a location of interest.

## Example in the book {-}

Modelling fine particulate matter PM2.5 in the USA in relation to temperature and precipitation.

$$Y_i \sim \mathcal{N}(\mu_i, \sigma^2)$$

$$\mu_i = \beta_0 + \beta_1 \cdot T_i + \beta_2 \cdot P_i + S(x_i)$$

## Preparations before fitting with `inla()` {-}

- create the mesh
- define the SPDE model
- create an index set for the SPDE model
- construct the projection matrix A
- create a stack with the data for estimation and prediction

## Preparations before fitting with `inla()` {-}

- create the mesh

```{r eval=FALSE}
mesh <- inla.mesh.2d(loc = coo, max.edge = c(200, 500),
                     cutoff = 1)
```

## Preparations before fitting with `inla()` {-}

- define the SPDE model, after choosing the smoothness parameter $\nu$ (e.g. set $\nu = 1$, which means $\alpha = 2$ in a 2D plane)

```{r eval=FALSE}
spde <- inla.spde2.matern(mesh = mesh, alpha = 2, constr = TRUE)
```

- create an index set for the SPDE model

```{r eval=FALSE}
indexs <- inla.spde.make.index("s", spde$n.spde)
> str(indexs)
List of 3
 $ s      : int [1:3747] 1 2 3 4 5 6 7 8 9 10 ...
 $ s.group: int [1:3747] 1 1 1 1 1 1 1 1 1 1 ...
 $ s.repl : int [1:3747] 1 1 1 1 1 1 1 1 1 1 ...
```

## Preparations before fitting with `inla()` {-}

In the model formula, the SPDE model and the index set will be used in defining the random effect:

`f(s, model = spde)`

## Preparations before fitting with `inla()` {-}

- construct the projection matrix A

```{r eval=FALSE}
A <- inla.spde.make.A(mesh = mesh, loc = coo)
```

## Preparations before fitting with `inla()` {-}

- create a stack with the data for estimation and prediction

```{r eval=FALSE}
# stack for estimation stk.e
stk.e <- inla.stack(
  tag = "est",
  data = list(y = d$value),
  A = list(1, A),
  effects = list(
    data.frame(
      b0 = rep(1, nrow(A)),
      covtemp = d$covtemp,
      covprec = d$covprec
    ),
    s = indexs
  )
)
# stack for prediction stk.p
stk.p <- inla.stack(
  tag = "pred",
  data = list(y = NA),
  A = list(1, Ap),
  effects = list(
    data.frame(
      b0 = rep(1, nrow(Ap)),
      covtemp = dp$covtemp,
      covprec = dp$covprec
    ),
    s = indexs
  )
)
# stk.full has stk.e and stk.p
stk.full <- inla.stack(stk.e, stk.p)
```

## Fit the model {-}

```{r eval=FALSE}
formula <- y ~ 0 + b0 + covtemp + covprec + f(s, model = spde)
res <- inla(
  formula,
  family = "gaussian",
  data = inla.stack.data(stk.full),
  control.predictor = list(
    compute = TRUE,
    A = inla.stack.A(stk.full)
  ),
  control.compute = list(return.marginals.predictor = TRUE)
)
```


## Extract results  {-}

- `res$summary.fixed`
- `res$summary.fitted.values`
- `res$marginals.fitted.values`

Extract needed subsets with: `inla.stack.index(stack = , tag =)`

Compute the probability of exceeding a threshold (PM2.5 >= 10), by using `inla.pmarginal()`:

`1 - inla.pmarginal(q = 10, marginal = marg)`

## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
