# Spatial point processes and simulation

**Learning objectives:**

- understand the difference between spatial point _patterns_ vs. _processes_
- learn about the properties of a Poisson process
- simulate Poisson processes with constant or with spatially varying intensity

## Spatial point process vs. spatial point pattern {-}

- **Spatial point process**: a _stochastic_ process $\{X_1, X_2, ..., X_{N(A)}\}$ taking values in a planar region $A \subset \mathbb{R}^2$.

  !! $N(A)$ is a random variable !!

- **Spatial point pattern**: a single _realization_ of a spatial point process.

  Each realization has a different number of points and different locations.

  The points are also often referred to as _events_.

## Intensity function of a spatial point process {-}

Intensity of a **spatial point pattern**: number of events per unit area.

$$\lambda = \frac{N(A)}{|A|}$$

For a **spatial point process**, $N(A)$ is a random variable, so its expected value is used to define the intensity parameter $\lambda$:

$$\lambda = \frac{E[N(A)]}{|A|}$$

## Intensity function of a spatial point process {-}

Consider the planar region $A$ as a position range represented by a variable $x$.

For a spatial point process we can define an intensity function that depends on $x$:

$$\lambda(x) = \frac{E[N(x)]}{|x|}$$

So $\lambda(x)$ now varies through space.

## Intensity function of a spatial point process {-}

We get the intensity for the study area by integration:

$$\lambda = \frac{E[N(A)]}{|A|} = \frac{\int_A{\lambda(x) dx}}{|A|}$$

## Commonly used special cases {-}

- **stationary**, **isotropic** spatial point process: has a spatially constant intensity function in any subregion $A_i$ of $A$:

$$\lambda(x) = \lambda = \frac{E[N(A)]}{|A|} = \frac{E[N(A_i)]}{|A_i|}$$

## Commonly used special cases {-}

- **Poisson** spatial point process:

    in any subregion $A_i$ of $A$:

1. the **number** of events follows a Poisson distribution:

$$N(A_i) \sim \mathcal{Poisson}(\mu_{A_i})$$

With:

$$\mu_{A_i} = E[N(A_i)] = \int_{A_i}{\lambda(x) dx}$$

## Commonly used special cases {-}

2. the **locations** of realized events are obtained as a random sample, with the inclusion probability in $A_i$ proportional to the intensity function $\lambda(x)$.

The general case, with spatially varying $\lambda(x)$, is called a **heterogeneous Poisson process**.

With spatially constant $\lambda(x) = \lambda$, we have a **homogeneous Poisson process**.

$$\mu_{A_i} = E[N(A_i)] = \int_{A_i}{\lambda(x) dx} = \lambda \cdot |A_i|$$

This is also called **CSR = complete spatial randomness**, which is tied to homogeneous spatial point processes.

## Poisson processes: what's next {-}

We can:

- simulate spatial point patterns from a homogeneous or heterogeneous Poisson process: **this chapter**
- test whether a given spatial point pattern deviates from CSR: **next chapter**

## Simulating spatial point patterns {-}

Simulation is useful e.g. to generate a distribution of process properties and compare it with:

- data (hypothesis testing for data)
- or some preset requirement (power analysis with simulated samples).

```{r message=FALSE}
library(spatstat)
```


```{r include=FALSE}
set.seed(20240808)
```


## Simulating spatial point patterns {-}

Create a **single realization** of a Poisson process with intensity $\lambda(x)$ in a specific observation window with:

```r
rpoispp(lambda = , win = )
```

The argument `lambda` is:

- a decimal number for _homogeneous_ Poisson processes
- a function for _heterogeneous_ Poisson processes

## Example: homogeneous {-}

```{r}
our_lambda <- 100
our_window <- owin(xrange = c(0, 1), yrange = c(0, 2))
```

The size of the observation window is 2 square units, so the expected number of events is 2 * 100 = 200 events.

```{r}
simulated_hom <- rpoispp(lambda = our_lambda, win = our_window)
simulated_hom
```

## Example: homogeneous {-}

```{r out.width="100%"}
plot(simulated_hom)
```

## Example: homogeneous {-}

How many points are there?

```{r}
npoints(simulated_hom)
```

Note:

```{r}
npoints(simulated_hom) == our_lambda * 2
```

Each realization has its own number of events, since this number is sampled from the Poisson distribution!

## Example: homogeneous {-}

Compare further realizations with an intensity set at 100:

```{r collapse=TRUE}
rpoispp(lambda = our_lambda, win = our_window) |> npoints()
rpoispp(lambda = our_lambda, win = our_window) |> npoints()
rpoispp(lambda = our_lambda, win = our_window) |> npoints()
rpoispp(lambda = our_lambda, win = our_window) |> npoints()
```

## Example: heterogeneous {-}

```{r}
our_lambda_fun <- function(x, y) {10 + 100 * x + 200 * y}
our_window <- owin(xrange = c(0, 1), yrange = c(0, 2))
```

## Example: heterogeneous {-}

What is the expected number of events in this observation window?

$$E[N([0,1] \times [0,2])] = \int_{[0,1] \times [0,2]}{\lambda(x, y) dx dy} = 520$$

## Example: heterogeneous {-}

Let's first have a look at the spatial variation of the intensity!

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Example: heterogeneous {-}

```{r intensity-gradient, eval=FALSE}
crossing(
  x = seq(0, 1, 0.02),
  y = seq(0, 2, 0.02)
) |>
  mutate(intensity = our_lambda_fun(x, y)) |>
  ggplot(aes(x = x, y = y, fill = intensity)) +
  geom_tile() +
  scale_fill_viridis_c() +
  coord_fixed() +
  theme_minimal() +
  ggtitle("Chosen spatial\nintensity gradient")
```


## Example: heterogeneous {-}

```{r ref.label="intensity-gradient", echo=FALSE, out.width="100%"}
```

## Example: heterogeneous {-}

```{r}
simulated_het <- rpoispp(lambda = our_lambda_fun, win = our_window)
simulated_het
```

## Example: heterogeneous {-}

```{r out.width="100%"}
plot(simulated_het)
```

## Example: heterogeneous {-}

```{r collapse=TRUE}
rpoispp(lambda = our_lambda_fun, win = our_window) |> npoints()
rpoispp(lambda = our_lambda_fun, win = our_window) |> npoints()
rpoispp(lambda = our_lambda_fun, win = our_window) |> npoints()
rpoispp(lambda = our_lambda_fun, win = our_window) |> npoints()
```


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
