# Point process modeling

```{r include=FALSE}
oldopt <- options(pillar.print_max = 5, pillar.print_min = 5)
```

**Learning objectives:**

- understand that spatial point process intensity can be modelled with a Gaussian Random Field (GRF)
- use R-INLA with spatial correlation to model intensity

## Considering intensity as a stochastic variable {-}

Up to now, we have considered intensity as a fixed parameter of either:

- the whole study region $A$ (homogeneous poisson process):

$$\lambda = \frac{E[N(A)]}{|A|}$$

- the location $x$ (inhomogeneous poisson process):

$$\lambda(x) = \lim\limits_{|dx| \to 0}\frac{E[N(dx)]}{|dx|}$$

## Considering intensity as a stochastic variable {-}

However we can approach intensity as a (locally) stochastic variable with a Poisson distribution:

$$\Lambda(s) \sim \mathcal{Poisson}(\mu_s)$$

To fit this model, we use the number of events in subregions $A_i$ of $A$ and use the area of $A_i$ as an offset:

$$|A_i| \cdot \Lambda(s)_{A_i} \sim \mathcal{Poisson}(|A_i| \cdot \mu_s)$$

## Considering intensity as a stochastic variable {-}

$\mu_s$ is fit using a log link and a linear predictor $\eta_s$:

$log(\mu_s) = \eta_s =$ fixed effects + spatial random effect (GRF) + unstructured random effect

A model (process) where a response variable can be expressed as a Gaussian process using a log link, is called a **Log-Gaussian Cox Process**.

## Considering intensity as a stochastic variable {-}

The GRF can be fitted using either a regular grid or a triangulated mesh.

- The book elaborates the latter, using INLA!
- The triangulated mesh approach doesn't rely on binning, but respects the exact location of events.

## Similarities with chapter 15 (model-based geostatistics) {-}

- Use of R-INLA
- Fitting models with a Gaussian Markov Random Field (GMRF) as a spatially correlated random effect

## Differences with chapter 15 (model-based geostatistics) {-}

- In geostatistical data, the stochastic process $Z(s)$ (a continuous response variable) can be observed everywhere in domain $D$.
- But in a spatial point process, domain $D$ is defined as where the events occur; so the domain itself is considered stochastic.
  - Point patterns arise when the variable to be analyzed corresponds to the _location of events_.

## Aim in chapter 15 (model-based geostatistics) {-}

Fit a statistical model that:

- can support various distribution families for the _response variable_
- accommodates fixed and random effects
- captures spatial correlation structure as a random effect
- provides spatial predictions with uncertainty measures

Specify the spatial random effect as a Gaussian Random Field (GRF) with zero mean and a Matérn correlation.

## How to model a spatial point process with INLA? {-}

- Model the **intensity** of the spatial point process over the whole study region.
So consider the intensity as the response variable.
- The linear predictor for the intensity is modelled as a local intercept plus a **GMRF with zero mean and a Matérn correlation** (spatial random effect).
- To do that, an **SPDE** (stochastic partial differential equation) is fitted on the vertices of a triangulated mesh.

## How to model a spatial point process with INLA? {-}

- We fit the intensity by modelling the number of points _per unit area_ as a **Poisson distribution**.
- The **area offset** for the mesh vertices (integration points) is determined by a **dual mesh**: polygons around the primary mesh vertices.
In effect, the offset serves as a weight.
  - The response at the mesh vertices is set to an initial observation of 0 points and offset according to the dual mesh.
  - The response at the event locations is set as 1 with an offset of 0.

## Projection matrix for spatial point processes {-}

As in chapter 15: 

GMRF values fitted at vertices are then interpolated to locations of interest:

- fitted values at observation locations
- predicted values at prediction locations

These interpolations need a **projection matrix**: defines the relation between locations and mesh vertices by means of weights.

## Projection matrix for spatial point processes {-}

As in chapter 15: 

- the rows represent the locations of interest (either observations or predictions).
- the columns represent the mesh vertices.
- the values are the barycentric coordinates of the locations of interest relative to the three vertices of the triangle, the latter having a mass of 1.
- So the rowsums are 1.

## Projection matrix for spatial point processes {-}

The values in the projection matrix are used as _weights_ to interpolate from observations to triangle vertices, or from triangle vertices to a prediction location.

For a spatial point process this is not different.
A projection matrix is created for:

- the **observation locations**: both the mesh vertices and the event locations (see before)
- the **prediction locations**

## Example in the book {-}

```{r message=FALSE}
library(sf)
library(terra)
library(tidyterra)
library(dplyr)
library(rnaturalearth)
library(ggplot2)
library(INLA)
```

```{r}
projUTM <- "+proj=utm +zone=19 +south +ellps=GRS80 \
            +towgs84=0,0,0,0,0,0,0 +units=km +no_defs"
```


## Example in the book {-}

Modelling the occurrence of the plant genus _Solanum_ in Bolivia between 2015 and 2022 (data from [GBIF](https://gbif.org)).

Using only a spatial GRF and a local intercept.

```{r}
d <- readr::read_csv("data/solanum.csv", col_types = "cddcDc") |> 
  select(longitude, latitude) |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |> 
  st_transform(projUTM)
map_4326 <- ne_countries(
  type = "countries",
  country = "Bolivia",
  scale = "medium",
  returnclass = "sf"
)
map <- map_4326 |> st_transform(projUTM)
```

## Example in the book {-}

```{r}
# modern way to do this conversion (but currently looses the CRS attribute):
map_4326 |> 
  st_geometry() |> 
  st_transform("EPSG:5356") |> 
  st_transform(pipeline = "+proj=unitconvert +xy_out=km")
```

## Example in the book {-}

Observed occurrences (spatial point pattern):

```{r}
d
coo <- st_coordinates(d)
(n <- nrow(coo))
```

## Example in the book {-}

```{r out.width="100%"}
ggplot() + 
  geom_sf(data = map, fill = "grey30") +
  geom_sf(data = d, colour = "pink", shape = 21) + 
  coord_sf(datum = projUTM) +
  theme_linedraw()
```

## Preparations before fitting with `inla()` {-}

Similar to chapter 15: 

- create the **mesh** and the **dual mesh**
- calculate the **offsets** (surface areas) to fit intensities instead of numbers
- define the **SPDE** model
- construct the **projection matrix `A.obs`** for observation locations
- construct the **projection matrix `A.p`** for prediction locations
- create a **stack** with the data for estimation and prediction

## Create the mesh {-}

```{r}
mesh <- inla.mesh.2d(
  loc.domain = st_coordinates(map)[, c("X", "Y")],
  max.edge = c(50, 100),
  offset = c(50, 100),
  cutoff = 1
)
(nmesh <- mesh$n)
```

## Create the mesh {-}

```{r out.width="100%", echo=FALSE}
plot(mesh, edge.color = "grey70", draw.segments = FALSE)
points(coo, pch = 20, cex = 0.25, col = "purple")
lines(st_coordinates(map)[, "X"], st_coordinates(map)[, "Y"])
```

## Create the dual mesh {-}

```{r}
source("R/book.mesh.dual_function.R")
dmesh <- book.mesh.dual(mesh)
```

## Create the dual mesh {-}

```{r out.width="100%", echo=FALSE}
plot(dmesh, border = "lightblue")
plot(mesh, add =  TRUE, edge.color = "grey70", draw.segments = FALSE)
```

## Calculate the offsets (surface areas) {-}

This is needed to fit intensities instead of numbers.
The surface areas serve as weights for the mesh vertices.

```{r}
dmesh_sf <- st_as_sf(dmesh)
st_crs(dmesh_sf) <- st_crs(map)
dmesh_sf <- dmesh_sf |> 
  mutate(id = row_number()) |> 
  st_sf(agr = "identity")

w <- st_intersection(dmesh_sf, st_geometry(map)) |>
  (\(x) mutate(x, area = st_area(x) |> as.numeric()))() |>
  st_drop_geometry() |>
  left_join(x = dmesh_sf, y = _, join_by(id)) |>
  mutate(area = ifelse(is.na(area), 0, area)) |>
  pull(area)
```

## Calculate the offsets (surface areas) {-}

`w` contains surface areas in km² within Bolivia:

```{r}
summary(w)
sum(w)
st_area(map)
```

## Define the SPDE model {-}

The smoothness parameter was chosen as $\nu = 1$, which means $\alpha = 2$ in a 2D plane.

```{r}
spde <- inla.spde2.matern(mesh = mesh, alpha = 2, constr = TRUE)
```

## Define the SPDE model {-}

In the model formula, the SPDE model will be used in defining the random effect:

`f(s, model = spde)`

## Construct the projection matrices {-}

Projection matrix for observation locations:

```{r}
# event locations
A.y <- inla.spde.make.A(mesh = mesh, loc = coo)
# mesh vertices
A.mesh <- Diagonal(nmesh, rep(1, nmesh))
# all observations
A.obs <- rbind(A.mesh, A.y)
dim(A.obs)
nmesh
nmesh + n
```

## Construct the projection matrices {-}

Construct prediction locations:

```{r}
grid <- rast(map, nrows = 100, ncols = 100)
# transform grid to a sf points object
dp <- grid |> 
  crds() |> 
  as.data.frame() |> 
  st_as_sf(coords = c("x", "y"), crs = st_crs(map))
# indices points within the map
indicespointswithin <- which(st_intersects(dp, map, sparse = FALSE))
# points within the map
dp <- st_filter(dp, map)
coop <- st_coordinates(dp)
```

## Construct the projection matrices {-}

```{r out.width="100%"}
ggplot() + 
  geom_sf(data = map, fill = "grey30") +
  geom_sf(data = dp, colour = "lightblue", size = 0.2) + 
  coord_sf(datum = projUTM) +
  theme_linedraw()
```

## Preparations before fitting with `inla()` {-}

Projection matrix for the prediction locations:

```{r}
A.p <- inla.spde.make.A(mesh = mesh, loc = coop)
```

## Stack with data for estimation and prediction {-}

Defining the observations & their offset:

```{r}
y.obs <- rep(0:1, c(nmesh, n))
e.obs <- c(w, rep(0, n)) # areas!!
```

Stack for estimation:

```{r}
stk.e <- inla.stack(
  tag = "est",
  data = list(y = y.obs, e = e.obs),
  A = list(1, A.obs),
  effects = list(
    list(b0 = rep(1, nmesh + n)),
    list(s = 1:nmesh)
  )
)
```

## Stack with data for estimation and prediction {-}

Stack for prediction:

```{r}
stk.p <- inla.stack(
  tag = "pred",
  data = list(y = rep(NA, nrow(coop)), e = rep(0, nrow(coop))),
  A = list(1, A.p),
  effects = list(
    data.frame(b0 = rep(1, nrow(coop))),
    list(s = 1:nmesh)
  )
)
```

Combine both stacks:

```{r}
stk.full <- inla.stack(stk.e, stk.p)
```

## Fit the model {-}

```{r}
formula <- y ~ 0 + b0 + f(s, model = spde)
```

```{r warning=FALSE, message=FALSE}
res <- inla(
  formula,
  family = 'poisson',
  data = inla.stack.data(stk.full),
  control.predictor = list(
    compute = TRUE,
    link = 1,
    A = inla.stack.A(stk.full)
  ),
  E = inla.stack.data(stk.full)$e
)
```

E is the offset: _the known component in the mean for the Poisson likelihoods, defined as $E_i \cdot e^{\eta_i}$_.

## Extract results  {-}

Extract needed subsets with: `inla.stack.index(stack = , tag =)`

```{r}
index <- inla.stack.index(stk.full, tag = "pred")$data
pred_mean <- res$summary.fitted.values[index, "mean"]
pred_ll <- res$summary.fitted.values[index, "0.025quant"]
pred_ul <- res$summary.fitted.values[index, "0.975quant"]

grid$ll <- NA
grid$mean <- NA
grid$ul <- NA

grid$mean[indicespointswithin] <- pred_mean
grid$ll[indicespointswithin] <- pred_ll
grid$ul[indicespointswithin] <- pred_ul
```

## Extract results  {-}

```{r}
grid
```

## Extract results  {-}

```{r prediction-plot, eval=FALSE}
ggplot() +
  geom_spatraster(data = grid) +
  coord_sf(datum = st_crs(grid)) +
  facet_wrap(~lyr) +
  scale_fill_viridis_c(na.value = "transparent", direction = -1) +
  theme_minimal()
```

## Extract results  {-}

```{r ref.label="prediction-plot", echo=FALSE, out.width="100%"}
```

```{r include=FALSE}
options(oldopt)
```

## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
